Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-04-11 endofquarterpaper.qmd
stash

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

@@

time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"))

trans_site_count <- nrow(trans_unique_data)
trans_sign_count <- ncol(trans_artifact_data)
proto_site_count <- nrow(proto_unique_data)
proto_sign_count <- ncol(proto_artifact_data)
early_site_count <- nrow(early_unique_data)
early_sign_count <- ncol(early_artifact_data)
evolved_site_count <- nrow(evolved_unique_data)
evolved_sign_count <- ncol(evolved_artifact_data)

nvogel <- nrow(signbase_full_clean %>% 
                 filter(site_name == "Vogelherd"))

@@

artifact_count_median <- median(artifact_count_df$artifact_count)

@@

## mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist( method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  # BM: Ensure binary data as input to jaccard distance
  mutate_all(~ as.numeric(. > 0)) %>% 
  vegdist("jaccard")

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

@@

The distribution of objects across sites is highly skewed (@fig-site-map C), with a median number of `r artifact_count_median` objects, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity. We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths. Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

These results are consistent with existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992]. Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of intergroup social networks at around 30 k BP, however, our network analysis shows evidence of intergroup social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted. Dividing the data up into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024]. Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].

Our work also adds to a growing awareness of diversity within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006] and de-emphasize diversity. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Belfer-Cohen_Goring-Morris_2002; @Maher_Conkey_2019; @Vanhaeren_d; @belfer-cohen2003; @bordes1970; @Shemer_Barzilai_Marder_2024; @Schwendler_2012; @bordes1963]. This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries.

One limitation of the study is the inclusion of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups. It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange. Ultimately, we chose to include them as evidence of cultural connection, as a greater number of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. However, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. However, these patterns should not be viewed as representing discrete, consistent groups to which all the sites within the group belong, but rather as representing opposite ends of a cultural variant spectrum. We adopt the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.

@@

The observed social dynamics of the phases of the Aurignacian technocomplex correspond with the existing data on the environmental and social conditions of the Upper Paleolithic. This tells us that the usage of artwork as a proxy for cultural difference is an effective method for understanding the social dynamics of prehistoric humans.

>>>>

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans. Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian. 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

@@

time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"))
```

```{r}
trans_site_count <- nrow(trans_unique_data)
trans_sign_count <- ncol(trans_artifact_data)
proto_site_count <- nrow(proto_unique_data)
proto_sign_count <- ncol(proto_artifact_data)
early_site_count <- nrow(early_unique_data)
early_sign_count <- ncol(early_artifact_data)
evolved_site_count <- nrow(evolved_unique_data)
evolved_sign_count <- ncol(evolved_artifact_data)

@@

sign_material_df <- signbase_full_clean %>% 
  group_by(material) %>% 
  summarise(count = n())%>% 
  arrange(desc(count))

iv_count <- sign_material_df %>% 
  filter(material == "ivory") %>% 
  pull(count)

bone_count <-sign_material_df %>% 
  filter(material == "bone") %>% 
  pull(count)

lime_count <- sign_material_df %>% 
  filter(material == "limestone") %>% 
  pull(count)

ant_count <- sign_material_df %>% 
  filter(material == "antler") %>% 
  pull(count)

tooth_count <- sign_material_df %>% 
  filter(material == "animal tooth") %>% 
  pull(count)

ochre_count <- sign_material_df %>% 
  filter(material == "ochre") %>% 
  pull(count)

undet_count <- sign_material_df %>% 
  filter(material == "undetermined") %>% 
  pull(count)

stone_row <- tibble("material" = c("stone"), "count" = c(1))

sign_material_df <- sign_material_df %>% 
  rbind(stone_row) %>% 
  filter(material != "stone indet.")

misc_string <- sign_material_df %>% 
  filter(count == 1) %>% 
  dplyr::select(material) %>% 
  c()

misc_string <- misc_string$material

object_type_df <- signbase_full_clean %>% 
  group_by(object_type) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
```

@@

## mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist(method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  mutate_all(~ as.numeric(. > 0)) %>% 
  vegdist("jaccard")

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map). `r iv_count` of the objects are made of ivory, `r bone_count` objects are made of bone, `r lime_count` objects are made of limestone, `r ant_count` objects are made of antler, `r tooth_count` objects are made of animal teeth, `r ochre_count` objects are made of ochre, and `r undet_count` objects are made of undetermined material. The rest of the objects are made of `r combine_words(misc_string)`, with one made of each material.

@@

The distribution of objects across sites is highly skewed, as shown in @fig-site-map C. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types in order to make sure that the results would not be significantly impacted by the exlcusion of object abundance. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase using Mantel tests. The results as shown in @tbl-mantel-table show that none of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm. Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

These results are consistent with existing data on the environmental and population change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].

Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 place the emergence of inter-group social networks at around 30 k BP, however, our network analysis shows evidence of inter-group social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted. Dividing the data into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024]. Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019]

One possible limitation is the potential ambiguity that arises with some of the sign types. While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements. We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. Potential ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains information on the specific order and patterns in which signs occurred.

@@

This study both adds to our understanding of Aurignacian social dynamics and suggests potential new directions for future research. The observed social dynamics produced by our methods align with the existing data on the environmental and population conditions of the Upper Paleolithic. We feel that our research opens up new opportunities for the application of these methods to data on artwork variation in other regions and time periods. Our research also suggests an earlier emergence of dynamic inter-connectivity than previously predicted, as we see weaker inter-connections beginning to emerge by 41 k BP, and stronger inter-connections by 38 k BP.  Future research directions might involve the analysis of environmental conditions and their interactions with social dynamics, in order to determine whether some of the sites that we have pointed to as potential aggregation sites function as such. Another potential future research direction would be the usage of the more specific data released in SignBase 2.0 in order to counter potential sign type ambiguity.
